Towards Proof-Reconstruction of Problems in Classical Propositional Logic

### 2017-10-13 paper/sections/proof-reconstruction.tex
[ paper ] introduction added to emulate rules section.

In the rest of this section, we provide a formal description using
type theory to build definitions and theorems for functions
necessary to emulate \Metis' inference rules. For this description,
we will write the type judgments with ``$a : A$'' to denote that $a$
has $\abbre{A}$ type. Types are written using small caps.
We use \Prop type for propositions and a the dependent type, $Γ ⊢ φ$ for theorems. In the latter type, $Γ$ has $\abbre{List}~\Prop$ type
and $φ$ has type $\Prop$. $\abbre{List}$ is the usual inductive type
for lists. Finally, we will denote by $φ ≡ ψ$, the syntactic
equivalence between propositions.

@@

From the definition above, the size of the argument, ((φ₁ ∧ φ₂) ⇒ φ₃),
is not structurally smaller than the input formula, (φ₁ ⇒ (φ₂ ⇒ φ₃)),
since ((φ₁ ∧ φ₂) ⇒ φ₃) is not a subformula of (φ₁ ⇒ (φ₂ ⇒ φ₃)).

@@

Finally, following the technique mentioned, we get a curated version of
$unshunt_{0}$ function presented in Eq.~(\ref{eq:unshunt-fixed}).

\begin{definition}[unshunt]
\begin{equation}
\label{eq:unshunt-fixed}
unshunt(\varphi) := unshunt_{1}(\varphi, unshunt_{cm}(\varphi)).
\end{equation}
\end{definition}

@@

@@

\begin{proof} Use induction on the cases defined by the outcome of the
$unshunt$ function in Eq.~(\ref{eq:unshunt-structured}).
\begin{itemize}
	\item When $n = 0$, by definition, $Γ ⊢ φ$.
	\item When $n = 1$, we apply the following theorem from~\cite{AgdaProp}
		\begin{equation*}
		\texttt{∧⇒-to-⇒⇒}\ :\  Γ ⊢ (φ₁ ∧ φ₂) ⇒ φ₃ → Γ ⊢ φ₁ ⇒ (φ₂ ⇒ φ₃).
		\end{equation*}
  \item For $n > 1$, we will use induction on the structure of the second
			  argument.

@@

\begin{equation}
\label{eq:split-definition}
\mysplit(φ) =
\begin{cases}
unshunt(\mysplit(φ₁)) &\text{if }φ ≡ φ₁ ∧ φ₂\\
∧\ unshunt(φ₁ ⇒ \mysplit(φ₂)), &\\[2.5mm]
unshunt(¬ φ₁ ⇒ \mysplit(φ₂)),
  & \text{if }φ ≡ φ₁ ∨ φ₂\\[2.5mm]
unshunt(φ₁ ⇒ \mysplit(φ₂)),
  & \text{if }φ ≡ φ₁ ⇒ φ₂\\[2.5mm]
unshunt(φ₁ ⇒ \mysplit(φ₂))
  & \text{if }φ ≡ φ₁ ⇔ φ₂\\
∧\ unshunt(φ₂ ⇒ \mysplit(φ₁)),
  &\\[2.5mm]
unshunt(φ₁ ⇒ \mysplit(¬ φ₂)),
  & \text{if }φ ≡ ¬ (φ₁ ∧ φ₂)\\[2.5mm]
unshunt(\mysplit(¬ φ₁))
  & \text{if }φ ≡ ¬ (φ₁ ∨ φ₂)\\
∧\ unshunt(¬ φ₁ ⇒ \mysplit(¬ φ₂)),
  &\\[2.5mm]
unshunt(\mysplit(φ₁))
  & \text{if }φ ≡ ¬ (φ₁ ⇒ φ₂)\\
∧\ unshunt(φ₁ ⇒ \mysplit(¬ φ₂)),
  &\\[2.5mm]
unshunt(φ₁ ⇒ \mysplit(¬ φ₂))
  & \text{if }φ ≡ ¬ (φ₁ ⇔ φ₂)\\
∧\ unshunt(¬ φ₂ ⇒ \mysplit(φ₁)),
  &\\[2.5mm]
unshunt(\mysplit(φ₁)),
  & \text{if }φ ≡ ¬ ¬ φ₁\\
⊤,
  & \text{if }φ ≡ ¬ ⊥\\
⊥,
  & \text{if }φ ≡ ¬ ⊤.
\end{cases}
\end{equation}
\end{definition}

@@

\begin{theorem}[thm-conjunct]
  \label{thm:thm-conjunct}
  $Γ ⊢ φ$, and $ψ  : \Prop$ , then $Γ ⊢ conjunct(φ, ψ)$.
\end{theorem}
~\begin{proof}\hspace{2cm}\\
\begin{itemize}
  \item For the case, $φ ≡ ψ$, $Γ ⊢ conjunct(φ, ψ)$ normalizes to $Γ ⊢ ψ$.
Then, we get the desire conclusion by applying the $subst$ theorem.
  \item If the proposition $φ$ is a conjunction, and we can get $ψ ≡ conjunct(φ_{i}, ψ)$ for some $i = 1,\ 2$, then,

@@

\begin{lemma}[lem-rdisj]
\label{lem:lem-rdisj}
  $Γ ⊢ φ$ then $Γ ⊢ rdisj(φ)$.
\end{lemma}
Following a similar way to define $rdisj$, we define the $rconj$ function,
to convert a conjunction into its right-associative form.

The $build_{∨}$ function in Eq.~(\ref{eq:build-or}) function intends to construct a disjunction from another disjunction, specifically, it rearranges the disjuncts in the source formula to match with the target disjunction formula as we described above.

@@

We define similar functions and state some theorems as we previously did for disjunctions, but instead, we treat conjunctive normal forms.

@@

\begin{proof}
\hspace{3mm}\\
\begin{itemize}
 \item
 \begin{equation*}
    \begin{bprooftree}
      \AxiomC{$Γ ⊢ φ₁$}
      \RightLabel{thm-reorder$_{∨}$.}
      \UnaryInfC{$Γ ⊢ reorder_{∨}(φ₁, l ∨ ψ)$}
   \end{bprooftree}
  \end{equation*}
\item
 \begin{equation*}
    \begin{bprooftree}
      \AxiomC{$Γ ⊢ φ₂$}
      \RightLabel{thm-reorder$_{∨}$.}
      \UnaryInfC{$Γ ⊢ reorder_{∨}(φ₂, ¬ l ∨ ψ)$}
   \end{bprooftree}
  \end{equation*}
  \item
  \begin{equation*}
  \begin{bprooftree}
    \AxiomC{$Γ ⊢ reorder_{∨}(φ₁, l ∨ ψ)$}
    \AxiomC{$Γ ⊢ reorder_{∨}(φ₂, ¬ l ∨ ψ)$}
    \RightLabel{∧-intro}
    \BinaryInfC{$Γ ⊢ reorder_{∨}(φ₁, l ∨ ψ) ∧ reorder_{∨}(φ₂, ¬ l ∨ ψ)$}
    \RightLabel{lem-rsol.}
    \UnaryInfC{$Γ ⊢ rsol(reorder_{∨}(φ₁, l ∨ ψ) ∧ reorder_{∨}(φ₂, ¬ l ∨ ψ))$}
  \end{bprooftree}
  \end{equation*}
\end{itemize}
\end{proof}

@@

% \subsubsection{Clausification.}
% \label{sssec:clausification}

% \begin{theorem}[thm-clausify]
% \label{thm:thm-clausify}
%   $Γ ⊢ φ$ and $ψ \in\Prop$ then $Γ ⊢ clausify(φ, ψ)$, where,
%   \begin{align}
%       clausify(φ,ψ) =
%       \begin{cases}
%         ψ, &\text{ if }φ≡ψ\\
%         reorder_{∧∨}(cnf(φ), ψ), &\text{otherwise.}
%       \end{cases}
%   \end{align}
% \end{theorem}

% \begin{proof} When $φ ≡ ψ$, we use the $subs$ theorem Eq.~(\ref{eq:substitution-theorem}). Otherwise, we use Theorem \ref{thm:thm-reorder-and-or} and Theorem \ref{thm:thm-cnf}.
% \end{proof}

% % subsubsection clausification (end)
% % --------------------------------------------------------------------

@@

% % -------------------------------------------------------------------
% \subsubsection{Simplify}
% \label{sssec:simplify}

% The \simplify rule is the application of the the lemma showed above,
% but also the following lemmas that intend to reduce a list of derivations,
% a conjunction of their formulas into a one smaller.

% \improvement{Missing theorems here.}

% % subsubsection simplify (end)
% % -------------------------------------------------------------------

>>>>

In the rest of this section, we provide a formal description using
type theory to build definitions and theorems for functions
necessary to emulate \Metis' inference rules. For this description,
we will write the type judgments with ``$a : A$'' to denote that $a$
has $\abbre{A}$ type. Types are written using small caps.
We use \Prop type for propositions and the dependent type, $Γ ⊢ φ$ for theorems. In the latter type, $Γ$ has $\abbre{List}~\Prop$ type
and $φ$ has type $\Prop$. $\abbre{List}$ is the usual inductive type
for lists. Finally, we will denote by $φ ≡ ψ$, the syntactic
equivalence between propositions.

@@

From the definition above, the size of the argument, $((φ₁ ∧ φ₂) ⇒ φ₃)$,
is not structurally smaller than the input formula, $(φ₁ ⇒ (φ₂ ⇒ φ₃))$,
since $((φ₁ ∧ φ₂) ⇒ φ₃)$ is not a subformula of $(φ₁ ⇒ (φ₂ ⇒ φ₃))$.

@@

Finally, following the technique mentioned, we get a curated version of
$unshunt_{0}$ function presented in the following definition.

\begin{definition}[unshunt]
\label{def:unshunt}
  \begin{align*}
    \begin{split}
    &unshunt : \Prop → \Prop\\
    &unshunt(\varphi) = unshunt_{1}(\varphi, unshunt_{cm}(\varphi)).
    \end{split}
  \end{align*}
\end{definition}

@@

We emulate each \Metis inference rule by proving a
function with the same name, and a theorem that
proves the soundness of the function.
The pattern is the following.

\begin{itemize}
\item \texttt{rule} is the \Metis inference rule.
\item $rule$ is the emulated version function for \texttt{rule} inference.
\begin{align*}
\begin{split}
&rule : \Prop → \Prop → \Prop\\
&rule(φ, ψ) =
  \begin{cases}
  ψ, &\text{if }rule\text{ built }ψ\text{ by applying valid transformations to }φ\\
  φ, &\text{otherwise.}
  \end{cases}
\end{split}
\end{align*}

\begin{itemize}
  \item $φ$ is the \emph{source} formula.
  \item $ψ$ is the \emph{target} formula.
\end{itemize}
\item thm-rule is the theorem, $Γ ⊢ rule(φ, ψ)$, that proves
the soundness of the $rule$ function.
\end{itemize}

@@

\begin{proof} Use induction on the cases defined by the outcome of the
$unshunt$ function in Eq.~(\ref{eq:unshunt-structured}).
\begin{itemize}
	\item When $n = 0$, by definition, $Γ ⊢ φ$.
	\item When $n = 1$, we apply the following theorem from~\cite{AgdaProp},
		\begin{equation*}
		\texttt{∧⇒-to-⇒⇒}\ :\  Γ ⊢ (φ₁ ∧ φ₂) ⇒ φ₃ → Γ ⊢ φ₁ ⇒ (φ₂ ⇒ φ₃).
		\end{equation*}
  \item For $n > 1$, we will use induction on the structure of the second
			  argument.

@@

\begin{equation}
\label{eq:split-definition}
\begin{split}
&split : \Prop → \Prop\\
&\mysplit(φ) =
\begin{cases}
unshunt(\mysplit(φ₁)) &\text{if }φ ≡ φ₁ ∧ φ₂\\
∧\ unshunt(φ₁ ⇒ \mysplit(φ₂)), &\\[2.5mm]
unshunt(¬ φ₁ ⇒ \mysplit(φ₂)),
  & \text{if }φ ≡ φ₁ ∨ φ₂\\[2.5mm]
unshunt(φ₁ ⇒ \mysplit(φ₂)),
  & \text{if }φ ≡ φ₁ ⇒ φ₂\\[2.5mm]
unshunt(φ₁ ⇒ \mysplit(φ₂))
  & \text{if }φ ≡ φ₁ ⇔ φ₂\\
∧\ unshunt(φ₂ ⇒ \mysplit(φ₁)),
  &\\[2.5mm]
unshunt(φ₁ ⇒ \mysplit(¬ φ₂)),
  & \text{if }φ ≡ ¬ (φ₁ ∧ φ₂)\\[2.5mm]
unshunt(\mysplit(¬ φ₁))
  & \text{if }φ ≡ ¬ (φ₁ ∨ φ₂)\\
∧\ unshunt(¬ φ₁ ⇒ \mysplit(¬ φ₂)),
  &\\[2.5mm]
unshunt(\mysplit(φ₁))
  & \text{if }φ ≡ ¬ (φ₁ ⇒ φ₂)\\
∧\ unshunt(φ₁ ⇒ \mysplit(¬ φ₂)),
  &\\[2.5mm]
unshunt(φ₁ ⇒ \mysplit(¬ φ₂))
  & \text{if }φ ≡ ¬ (φ₁ ⇔ φ₂)\\
∧\ unshunt(¬ φ₂ ⇒ \mysplit(φ₁)),
  &\\[2.5mm]
unshunt(\mysplit(φ₁)),
  & \text{if }φ ≡ ¬ ¬ φ₁\\
⊤,
  & \text{if }φ ≡ ¬ ⊥\\
⊥,
  & \text{if }φ ≡ ¬ ⊤.
\end{cases}
\end{split}
\end{equation}
\end{definition}

@@

\begin{theorem}[thm-conjunct]
  \label{thm:thm-conjunct}
  $Γ ⊢ φ$, and $ψ  : \Prop$ , then $Γ ⊢ conjunct(φ, ψ)$.
\end{theorem}
\begin{proof}\hspace{2cm}
\begin{itemize}
  \item For the case, $φ ≡ ψ$, $Γ ⊢ conjunct(φ, ψ)$ normalizes to $Γ ⊢ ψ$.
Then, we get the desire conclusion by applying the $subst$ theorem.
  \item If the proposition $φ$ is a conjunction, and we can get $ψ ≡ conjunct(φ_{i}, ψ)$ for some $i = 1,\ 2$, then,

@@

\begin{lemma}[lem-rdisj]
\label{lem:lem-rdisj}
  $Γ ⊢ φ$ then $Γ ⊢ rdisj(φ)$.
\end{lemma}
% Following a similar way to define $rdisj$, we define the $rconj$ function,
% to convert a conjunction into its right-associative form.

The $build_{∨}$ function in Eq.~(\ref{eq:build-or}) intends to construct a disjunction from another disjunction, specifically, this functions will try to rearrange the disjuncts in the source formula to match with the target disjunction formula, similar as we described above.

@@

We define similar functions and state some theorems as we previously did for disjunctions, but instead, we treat only with conjunctive normal forms.

@@

\begin{proof}\hspace{3mm}
\begin{itemize}
 \item
 \begin{equation*}
    \begin{bprooftree}
      \AxiomC{$Γ ⊢ φ₁$}
      \RightLabel{thm-reorder$_{∨}$.}
      \UnaryInfC{$Γ ⊢ reorder_{∨}(φ₁, l ∨ ψ)$}
   \end{bprooftree}
  \end{equation*}
\item
 \begin{equation*}
    \begin{bprooftree}
      \AxiomC{$Γ ⊢ φ₂$}
      \RightLabel{thm-reorder$_{∨}$.}
      \UnaryInfC{$Γ ⊢ reorder_{∨}(φ₂, ¬ l ∨ ψ)$}
   \end{bprooftree}
  \end{equation*}
  \item
  \begin{equation*}
  \begin{bprooftree}
    \AxiomC{$Γ ⊢ reorder_{∨}(φ₁, l ∨ ψ)$}
    \AxiomC{$Γ ⊢ reorder_{∨}(φ₂, ¬ l ∨ ψ)$}
    \RightLabel{∧-intro}
    \BinaryInfC{$Γ ⊢ reorder_{∨}(φ₁, l ∨ ψ) ∧ reorder_{∨}(φ₂, ¬ l ∨ ψ)$}
    \RightLabel{lem-rsol.}
    \UnaryInfC{$Γ ⊢ rsol(reorder_{∨}(φ₁, l ∨ ψ) ∧ reorder_{∨}(φ₂, ¬ l ∨ ψ))$}
  \end{bprooftree}
  \end{equation*}
\end{itemize}
\end{proof}

@@

% --------------------------------------------------------------------

\subsubsection{Clausification.}
\label{sssec:clausification}

The \clausify rule transforms a
propositional formula in its clausal normal form, a conjunction
of clauses. Where a \emph{clause} is the disjunction of zero or more
literals and a \emph{literal} is an atom (positive literal) or a
negation of an atom (negative literal). Be aware this kind of conversion between one formula to its clausal normal form is not unique, and \Metis has a customized approach to perform that transformation.
In the following \Metis \TSTP derivation, we saw a clear example, where
clausify expands the formula \texttt{normalize\_0\_0} by using distributive
laws.

\begin{verbatim}
fof(normalize_0_0, plain, (~ p | (q & r)),
    inference(canonicalize, [], [a1])).
fof(normalize_0_1, plain, ((~ p | q) & (~ p | r)),
    inference(clausify, [], [normalize_0_0])).
\end{verbatim}

\begin{definition}[clausify]
  \label{def:clausify}
 \begin{align*}
   \begin{split}
      &clausify : \Prop → \Prop → Prop\\
      &clausify(φ,ψ) =
      \begin{cases}
        ψ, &\text{ if }φ≡ψ\\
        reorder_{∧∨}(cnf(φ), ψ), &\text{otherwise.}
      \end{cases}
      \end{split}
  \end{align*}
\end{definition}

\begin{theorem}[thm-clausify]
\label{thm:thm-clausify}
  $Γ ⊢ φ$ and $ψ \in\Prop$ then $Γ ⊢ clausify(φ, ψ)$
\end{theorem}

\begin{proof} When $φ ≡ ψ$, we use the $subs$ theorem Eq.~(\ref{eq:substitution-theorem}). Otherwise, we use Theorem \ref{thm:thm-reorder-and-or} and Theorem \ref{thm:thm-cnf}.
\end{proof}

% subsubsection clausification (end)
% --------------------------------------------------------------------

@@

% -------------------------------------------------------------------
\subsubsection{Simplify}
\label{sssec:simplify}

The \simplify rule is an inference that performs simplification of
definitions and tautologies. This rule
transverses a list of formulas by applying different theorems (e.g.,
\emph{modus pones}, \emph{modus tollens}, or \emph{disjunctive
syllogism}) including those theorems for removing redundancies and definitions  from the description of the \canonicalize rule,
Fig. \ref{fig:conjunctive-disjunctive-simpl}. This rules also use the resolution as we show in the description of the \resolve rule~\ref{thm:thm-resolve}. Let see the following excerpt from a \Metis \TSTP derivation where \simplify appears.

\begin{verbatim}
fof(normalize_1_0, plain, (~ p & ~ r & q),
    inference(canonicalize, [], [negate_1_0])).
fof(normalize_1_1, plain, ((p | q) & (p | r)),
    inference(canonicalize, [], [a1])).
fof(normalize_1_2, plain, (p | r),
    inference(conjunct, [], [normalize_1_1])).
fof(normalize_1_3, plain, ($false),
    inference(simplify, [], [normalize_1_0, normalize_1_2])).
\end{verbatim}

% subsubsection simplify (end)
% -------------------------------------------------------------------
